---
title: "Visual Impact Assessment capability development - phase 3"
---

::: {.callout-warning}
The viewshed maps on this page are very detailed and may take quite a while to display in many web browsers, and may not render at all on small devices such as smartphones or tablets. They are provided purely for development purposes, and final versions will be optimised for faster display.
:::

## Calculating compound viewsheds

```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
#| echo: false
#| message: false
#| warning: false
#| include: false

library(tidyverse)
library(leaflet)
library(RColorBrewer)
library(terra)
library(terrainr)
library(raster)
library(osmdata)
library(sf)

terraOptions(progress=0)

if (!file.exists("_wip/oberon_and_environs_dem_5m.tif")) {

  # NSW Elevation Data Service
  # https://portal.spatial.nsw.gov.au/portal/apps/webappviewer/index.html?id=437c0697e6524d8ebf10ad0d915bc219
  
wallerawang_rast <- terra::rast("~/NSW_Government_Spatial_Services/Wallerawang-DEM-AHD_56_5m/Wallerawang-DEM-AHD_56_5m.asc")
orange_rast <- terra::rast("~/NSW_Government_Spatial_Services/Orange-DEM-AHD_55_5m/Orange-DEM-AHD_55_5m.asc")
bathurst_rast <- terra::rast("~/NSW_Government_Spatial_Services/Bathurst-DEM-AHD_55_5m/Bathurst-DEM-AHD_55_5m.asc")
blayney_rast <- terra::rast("~/NSW_Government_Spatial_Services/Blayney-DEM-AHD_55_5m/Blayney-DEM-AHD_55_5m.asc")
oberon_rast <- terra::rast("~/NSW_Government_Spatial_Services/Oberon-DEM-AHD_55_5m/Oberon-DEM-AHD_55_5m.asc")
crookwell_rast <- terra::rast("~/NSW_Government_Spatial_Services/Crookwell-DEM-AHD_55_5m/Crookwell-DEM-AHD_55_5m.asc")
katoomba_rast <- terra::rast("~/NSW_Government_Spatial_Services/Katoomba-DEM-AHD_56_5m/Katoomba-DEM-AHD_56_5m.asc")
taralga_rast <- terra::rast("~/NSW_Government_Spatial_Services/Taralga-DEM-AHD_55_5m/Taralga-DEM-AHD_55_5m.asc")
burragorang_rast <- terra::rast("~/NSW_Government_Spatial_Services/Burragorang-DEM-AHD_56_5m/Burragorang-DEM-AHD_56_5m.asc")

terra::writeRaster(wallerawang_rast, "_wip/wallerawang_dem.tif", overwrite=TRUE)
terra::writeRaster(orange_rast, "_wip/orange_dem.tif", overwrite=TRUE)
terra::writeRaster(bathurst_rast, "_wip/bathurst_dem.tif", overwrite=TRUE)
terra::writeRaster(blayney_rast, "_wip/blayney_dem.tif", overwrite=TRUE)
terra::writeRaster(oberon_rast, "_wip/oberon_dem.tif", overwrite=TRUE)
terra::writeRaster(crookwell_rast, "_wip/crookwell_dem.tif", overwrite=TRUE)
terra::writeRaster(katoomba_rast, "_wip/katoomba_dem.tif", overwrite=TRUE)
terra::writeRaster(taralga_rast, "_wip/taralga_dem.tif", overwrite=TRUE)
terra::writeRaster(burragorang_rast, "_wip/burragorang_dem.tif", overwrite=TRUE)

rm(wallerawang_rast)
rm(orange_rast)
rm(bathurst_rast)
rm(blayney_rast)
rm(oberon_rast)
rm(crookwell_rast)
rm(katoomba_rast)
rm(taralga_rast)
rm(burragorang_rast)

terrainr::merge_rasters(
  c("_wip/wallerawang_dem.tif",
    "_wip/orange_dem.tif",
    "_wip/bathurst_dem.tif",
    "_wip/blayney_dem.tif",
    "_wip/oberon_dem.tif",
    "_wip/crookwell_dem.tif",
    "_wip/katoomba_dem.tif",
    "_wip/taralga_dem.tif",
    "_wip/burragorang_dem.tif"),
   output_raster = "_wip/oberon_and_environs_dem_5m.tif",
  overwrite=TRUE)
}

localtif <- terra::rast("_wip/oberon_and_environs_dem_5m.tif")  

palings_yard_turbines_untransformed_sf <- sf::read_sf("assets/paling_yards_turbines.geojson")
palings_yard_turbines_sf <- st_transform(palings_yard_turbines_untransformed_sf, crs=crs(localtif))
palings_yard_turbines_df <- as.data.frame(st_coordinates(palings_yard_turbines_sf))

palings_yard_turbines_untransformed_df <- as.data.frame(st_coordinates(palings_yard_turbines_untransformed_sf)) %>%
  rename(long=X, lat = Y)

# Now create the summed viewsheds for all 47 turbines in Palings Yard
if (!file.exists("_wip/vshed_number_visible.tif")) {
  for (i in 1:nrow(palings_yard_turbines_untransformed_sf)) {
    print(i)
    vshed_raster <- viewshed(x=localtif,
                             loc=c(st_coordinates(palings_yard_turbines_sf[i,])), 
                             observer=240,
                             target=1.8,
                             curvcoef=0.85714,
                             output="yes/no")
    vshed_raster_10m <- terra::aggregate(vshed_raster, fact=2, fun=max)
    vshed_raster_10m <- terra::ifel(vshed_raster_10m <= 0, NA, vshed_raster_10m)
  
    # sum the number of towers visible for each cell in raster
    if (i == 1) {
      vshed_number_visible <- vshed_raster_10m
    } else {
      vshed_number_visible <- sum(vshed_number_visible,
                              vshed_raster_10m,
                              na.rm = TRUE)     
    }  
    rm(vshed_raster)
    rm(vshed_raster_10m)
    gc(verbose=TRUE, full=TRUE)
  
  }
  writeRaster(vshed_number_visible, "_wip/vshed_number_visible.tif")
} 

rm(localtif)

vshed_number_visible <- terra::rast("_wip/vshed_number_visible.tif") 

number_visible_pal <- colorNumeric(
  palette = "Reds",
  domain = 1:47,
  reverse = FALSE,
  na.color = "transparent")

# Other features
taralga_line <- readLines("assets/taralga_500kVA_line_v2.geojson") %>%
                  paste(collapse = "\n")

pines_expl_area <- sf::read_sf("assets/pines_exploration_area_v15.geojson")

oberon_boundary <- sf::read_sf("assets/nsw-lga-boundaries.geojson") %>%
                      filter(abb_name == "Oberon")

```

In this step, we will calculate the cumulative viewshed for all of the 47 turbines proposed for the Palings Yard wind farm, assuming each one is 240m tall. This assumptin will be varied in the next step.

:::{.column-page}

```{r, eval=FALSE}
#| echo: false
#| message: false
#| warning: false

leaflet(height="800px", width="1200px") %>% 
  setView(lng = palings_yard_turbines_untransformed_df[20,1],
          lat = palings_yard_turbines_untransformed_df[20,2],
          zoom = 10) %>%
  addProviderTiles("Esri.WorldShadedRelief", group="Shaded relief") %>%
  addProviderTiles("OpenTopoMap", group="Topography") %>%
  addTiles(group = "OpenStreetMap") %>%
  addProviderTiles("Esri.WorldImagery", group="Imagery") %>%
  addRasterImage(vshed_raster_1_10m, 
                 opacity=0.5, 
                 colors = vshed_raster_1_pal,
                  group = "Viewshed for 240m turbine 1",
                  maxBytes = 4 * 8330 * 8801) %>%
  addRasterImage(vshed_raster_2_10m, 
                 opacity=0.5, 
                 colors = vshed_raster_2_pal,
                  group = "Viewshed for 240m turbine 2",
                  maxBytes = 4 * 8330 * 8801) %>%
  addRasterImage(vshed_raster_3_10m, 
                 opacity=0.5, 
                 colors = vshed_raster_3_pal,
                  group = "Viewshed for 240m turbine 3",
                  maxBytes = 4 * 8330 * 8801) %>%
  addCircleMarkers(data = palings_yard_turbines_untransformed_df[c(2,20,45),],
              group = "Three proposed 240m turbines",
              color = "cyan",
              radius = 5,
              stroke = FALSE,
              fillOpacity = 0.5) %>%  
  addLayersControl(baseGroups = c("Imagery", "Shaded relief", "Topography", "OpenStreetMap"),
                   overlayGroups = c( "Three proposed 240m turbines",
                                      "Viewshed for 240m turbine 1",
                                      "Viewshed for 240m turbine 2",
                                      "Viewshed for 240m turbine 3"),
                   options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addMeasure(
            position = "bottomleft",
            primaryLengthUnit = "kilometers",
            primaryAreaUnit = "hectares",
            activeColor = "#3D535D",
            completedColor = "#7D4479",
            localization = "en") %>%
  addScaleBar(position = "bottomright",
  options = scaleBarOptions(metric=TRUE, 
                            imperial=FALSE,
                            maxWidth=300,
                            updateWhenIdle=TRUE))

                    
```

:::

We can examine the distribution of the number of turbines visible from each pixel in the cumulative viewshed raster. Note that zero turbines are visible from the vast majority of pixels -- these have been excluded from the chart below.

```{r, fig.asp=1.25}
#| echo: false
#| message: false
#| warning: false

vshed_number_visible_df <- as.data.frame(vshed_number_visible) %>%
                              mutate(number_visible = factor(viewshed))

vshed_number_visible_df  %>%
  ggplot(aes(x=number_visible)) +
    geom_bar(colour="orange", fill="orange") + 
    coord_flip() +
    labs(x="Number of turbines visible from each pixel in raster",
         y="Number of pixels",
         caption = "Note: Pixels from which no turbines are visible are excluded") +
    theme_minimal()
         
```

The map below shows the cumulative viewshed (with Earth curvature and atmospheric refraction corrections) for all 47 of the proposed Palings Yard wind turbines each 240 metres tall at the indicated positions, shaded by the number of turbines visible from each point on the map.

:::{.column-page}

```{r}
#| echo: false
#| message: false
#| warning: false

leaflet(height="800px", width="1200px") %>% 
  setView(lng = palings_yard_turbines_untransformed_df[20,1],
          lat = palings_yard_turbines_untransformed_df[20,2],
          zoom = 10) %>%
  addProviderTiles("Esri.WorldImagery", group="Imagery") %>%
  addProviderTiles("Esri.WorldShadedRelief", group="Shaded relief") %>%
  addProviderTiles("OpenTopoMap", group="Topography") %>%
  addTiles(group = "OpenStreetMap") %>%
  addRasterImage(vshed_number_visible, 
                 opacity=0.6, 
                 colors = number_visible_pal,
                  group = "Cumulative viewshed for all turbines",
                  maxBytes = 4 * 8330 * 8801) %>%
  addLegend(
    position = "bottomright",
    pal = number_visible_pal,
    values = 1:47,
    labels = palette(),
    title = "Number of turbines visible") %>%  
  addCircleMarkers(data = palings_yard_turbines_untransformed_df,
              group = "Proposed Palings Yard turbines",
              color = "cyan",
              radius = 5,
              stroke = FALSE,
              fillOpacity = 0.5) %>%  
  addPolygons(data=oberon_boundary,
              stroke=TRUE, 
              weight=4, 
              fillOpacity=0.1, 
              color="orange",
              group = "Oberon LGA") %>%
    addGeoJSON(taralga_line, 
             weight=3, 
             color="blue", 
             fillOpacity=0,
             group = "500 kVA transmission line (partial)") %>%
  addLayersControl(baseGroups = c("Imagery", "Shaded relief", "Topography", "OpenStreetMap"),
                   overlayGroups = c( "Oberon LGA",
                                      "500 kVA transmission line (partial)",
                                      "Proposed Palings Yard turbines",
                                      "Cumulative viewshed for all turbines"),
                   options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addMeasure(
            position = "bottomleft",
            primaryLengthUnit = "kilometers",
            primaryAreaUnit = "hectares",
            activeColor = "#3D535D",
            completedColor = "#7D4479",
            localization = "en")
```

:::


:::{.column-page}

```{r, eval=FALSE}
#| echo: false
#| message: false
#| warning: false

leaflet(height="800px", width="1200px") %>% 
  setView(lng = palings_yard_turbines_untransformed_df[20,1],
          lat = palings_yard_turbines_untransformed_df[20,2],
          zoom = 10) %>%
  addProviderTiles("Esri.WorldImagery", group="Imagery") %>%
  addProviderTiles("Esri.WorldShadedRelief", group="Shaded relief") %>%
  addProviderTiles("OpenTopoMap", group="Topography") %>%
  addTiles(group = "OpenStreetMap") %>%
  addRasterImage(vshed_number_visible_1, 
                 opacity=0.6, 
                 colors = "orange",
                 group = "One turbine visible (orange)",
                 maxBytes = 4 * 8330 * 8801) %>%
  addRasterImage(vshed_number_visible_2, 
                 opacity=0.6, 
                 colors = "red",
                 group = "Two turbines visible (red)",
                 maxBytes = 4 * 8330 * 8801) %>%
  addRasterImage(vshed_number_visible_3, 
                 opacity=0.6, 
                 colors = "darkred",
                 group = "Three turbines visible (dark red)",
                 maxBytes = 4 * 8330 * 8801) %>%
  addCircleMarkers(data = palings_yard_turbines_untransformed_df[c(2,20,45),],
              group = "Three proposed 240m turbines",
              color = "cyan",
              radius = 5,
              stroke = FALSE,
              fillOpacity = 0.5) %>%  
  addLayersControl(baseGroups = c( "Topography", "Shaded relief", "Imagery", "OpenStreetMap"),
                   overlayGroups = c( "Three proposed 240m turbines",
                                      "One turbine visible (orange)",
                                      "Two turbines visible (red)",
                                      "Three turbines visible (dark red)"),
                   options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addMeasure(
            position = "bottomleft",
            primaryLengthUnit = "kilometers",
            primaryAreaUnit = "hectares",
            activeColor = "#3D535D",
            completedColor = "#7D4479",
            localization = "en")
```

:::

## Next steps

The next step is to compute viewshed layers for each turbine  for different degrees of exposure of each turbine -- that is, whether only the tips of the upper blade arc can be seen, or the hub, or the entire blade arc. Plus a layer for the distance. Then these layers can be combined using various algorithms as described in [this paper](https://www.sciencedirect.com/science/article/pii/S016920462200072X), to give a visual impact metric for each turbine at each pixel location in the map. Then those can be summed (or aggregated by a function, not necessarily a linear sum) to give an overall visual impact metric for each pixel, which can then be displayed as binned layers on a single map. That's the full extent of what that paper mentioned above describes. That paper seems to be well-regarded and is a good starting point. And then finally drape that overall visual impact metric raster over a 3D rendered map with animated fly-arounds. Weighting of visual impact metrics by population density at each point is also possible to allow an approximate overall human population visual impact score to be created for a given configuration of wind turbines (and variations of that for specific locations or sets of locations). That would then allow the visual impact of different turbine layouts and numbers to be compared and optimised to reduce visual impact to a minimum (subject to other optimisation constraints).
